### 2021-05-14 submission/manuscript.Rmd
Add abstract, methods, discussion

## Abstract
(Modify depending on target journal, currently abstract submitted to World Microbe Forum) Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the intestinal microbiome. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630 spores. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge (dpc). To examine how PEG treatment impacts clearance, we administered PEG for 1 day to clindamycin-treated, *C. difficile*-challenged mice either immediately following challenge or 3 dpc. We collected longitudinal stool samples to examine *C. difficile* levels in the stool via anaerobic culture and profiled the microbiota by 16S rRNA sequencing. PEG treatment alone was sufficient to render mice susceptible to CDI and 5-day PEG-treated mice remain colonized for up to 30 dpc. Additionally, 5-day PEG treated mice remained susceptible to CDI 10-days post treatment. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpc. Although 5-day PEG-treated mice exhibited prolonged *C. difficile* colonization, we saw no difference in histological inflammation between PEG- and clindamycin-treated mice. Additionally, administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpc in mice that received PEG immediately after challenge and 15 dpc in mice that received PEG 3 dpc. When we examined microbiota composition across our different treatment groups, we found decreased richness in the PEG-treated mice that exhibited prolonged *C. difficile* colonization. Importantly, there were increased Bacteroides and Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in most of the PEG-treated mice with prolonged *C. difficile* colonization. Our findings suggest the osmotic laxative PEG 3350 alters the mouse microbiota and disrupts colonization resistance to *C. difficile*, as well as clearance in mice with a CDI. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact human microbiota colonization resistance.

@@

## Introduction
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because they disrupt microbiota colonization resistance [@Britton2014]. However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020]. Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].

@@

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives in the context of CDIs. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI and  than mice treated with clindamycin, and if PEG treatment after colonization can promote *C. difficile* clearance.

## Results
**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, 1. received no additional treatment, 2. was also treated with clindamycin, and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice relative to their baseline weights, with the greatest change in weight observed on the fifth day of PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). Therefore, PEG treatment led to sustained colonization in contrast to clindamycin mice that naturally cleared *C. difficile* within ten days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumable due to *C. difficile* as the bacteria became detecable in that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

```
**5-day laxative treatment differentially disrupted the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015]. We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

@@

Interestingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine if there were genera that changed during the post-challenge period when the group median *C. difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig. S1A). We found no bacteria were significantly different over the two time periods after multiple hypothesis correction (Data Set S1, Sheet X). However, there was also wide variation between individual mice regarding when *C. difficile* became detectable (Fig. S1A) as well as the relative abundances of bacterial genera in the communities (Fig. S1B). For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period and this corresponded to mouse 10, which died on the sixth day post-challenge and mouse 11, where *C. difficile* was present at a high density from the 4th day post-challenge onward (Fig. S1B). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

```
**5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha diversity continued to increase over time with the PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet X). Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C. difficile*) was one of the genera that had a significant difference in relative abundance between treatment groups at 6 days post-challenge. This population was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). By 30 days post-challenge, only the relative abundances of *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Next, we examined the severity of *C. difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5-2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

```
**Post-CDI laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. To examine the impact of PEG treatment on *C. difficile* clearance, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B). Next we examined how post-CDI PEG treatment impacted microbiota diversity. Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D).

@@

``` {r ml results}
#AUC results
auc_results <-  read_csv(here("results/performance_results.csv")) %>%
  mutate(taxa_level = "genus") %>%  
  group_by(method) %>%
  summarize(median = median(AUC)) %>%
  arrange(desc(median))
#Highest AUC value
max_auc <- auc_results %>% select(median) %>% max()

#To Do: Insert minimum Percent identity as variable name once blast results to Muribaculum have been combined into 1 file

```
**Five-day post-challenge community data can predict which mice that will have prolonged *C. difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C. difficile* colonization. We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C. difficile* 10 days post-challenge. We chose 5 days post-challenge because that was the earliest time point where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after challenge. The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C. difficile *colonization. We selected the top 10 bacteria contributing to our models performance (Fig. 7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C. difficile* colonization status on day 10 (Fig. 7B). Next, we focused on the 5 genera that had a greater than 1 % relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time. We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, the relative abundance of *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C. difficile* (Fig. 7C). Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-CDI PEG treated mice.

## Discussion
By following osmotic laxative treated mice over time, we found 5-day PEG treatment before challenge resulted in prolonged *C. difficile* colonization upon challenge, while a 1-day PEG treatment resulted in transient colonization. The differences in *C. difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota. Additionally, the intestinal communities of 5-day PEG treated mice had not regained *C. difficile* colonization resistance after a 10-day recovery period. In contrast to the hypothesis suggested by the literature, treating mice with PEG after *C. difficile* challenge prolonged colonization. We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C. difficile* colonization (Fig. 8). Overall, our results demonstrated that an osmotic laxative treatment alone rendered mice susceptible to *C. difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after *C. difficile* challenge.)

Some of the lower abundance bacteria could also play important roles in clearing C. difficile. For example, the relative abundance of Oscillibacter was increased in mice that cleared *C. difficile* compared to the mice with prolonged colonization. Oscillibacter __ can produce the SCFA valerate, which has been shown to inhibit *C. difficile* growth based on ___ data. Similarly, Acetatifactor can produce __, another SCFA with ___ role in the context of CDIs. Thus bacteria with potential protective functions might have played a role in promoting *C. difficile* clearance.

There was an interesting pattern of increased Enterobacteriaceae, Bacteroides, and C. difficile in both the stool and mucosal communities of PEG-treated mice suggesting a potential synergy. 
Akkermansia is also a mucin degrader with potentially beneficial or detetrimental roles depending on context in other diseases (). In our study the relative abundance of *Akkermansia* shifted over time between groups. In the stool it was initially increased in mice that cleared *C. difficile*, but shifted after 5-days post-challenge so that it was increased in mice that had prolonged colonization. In the context of CDIs, some studies suggest a protective role, while others suggest detrimental.

* Discussion of prolonged persistence. *C. difficile* sequences detected in tissue samples. Association with mucin-degrading bacteria suggested by recent papers. Akkermansia, Bacteroides are known mucin degraders and could be compete.

Bacteroides and Enterobacteriaceae synergy in the context of CRC (Dejea paper?)

* Protective bacteria missing in PEG-treated mice

>>>>

## Abstract
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge (dpc). PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remain colonized for up to 30 dpc. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpc. To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C. difficile*-challenged mice. Administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpc. When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90). Five of the top ten bacterial features important for predicting prolonged colonization had high relative abundances in the community. Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C. difficile* colonization in PEG-treated mice.

## Importance
Diarrheal samples induced by medications such as laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are some microbiota similarities between diarrheal and *C. difficile* infection (CDI) communities such as lower diversity compared to healthy controls, which lead us to hypothesize that diarrhea may be an indicator of *C. difficile* risk. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge Our findings suggest the osmotic laxative disrupts colonization resistance to *C. difficile*, as well as clearance in mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

## Introduction
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because they disrupt microbiota colonization resistance [@Britton2014]. However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020]. Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiota is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].

@@

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives in the context of CDIs. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI than mice treated with clindamycin, and if PEG treatment after challenge can promote *C. difficile* clearance.

## Results
**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days: 1. received no additional treatment, 2. was also treated with clindamycin, and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice relative to their baseline weights, with the greatest change in weight observed on the fifth day of PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). Therefore, PEG treatment led to sustained colonization in contrast to clindamycin mice that naturally cleared *C. difficile* within ten days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumably due to *C. difficile* as the bacteria became detectable in stool samples from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

```
**5-day laxative treatment differentially disrupted the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015]. We tried to minimize the impact of cage effects on our experiment by breaking up littermates when assigning mice to treatment groups. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group did no significantly explain the observed variation in fecal communities (Data Set S1, Sheet X). None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

@@

Interestingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine if there were genera that changed during the post-challenge period when the group median *C. difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig. S1A). We found no bacteria with relative abundances that were significantly different over the two time points after multiple hypothesis correction (Data Set S1, Sheet X). However, there was also wide variation between individual mice regarding when *C. difficile* became detectable (Fig. S1A) as well as the relative abundances of bacterial genera in the communities (Fig. S1B). For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period and this corresponded to mouse 10, which died on the sixth day post-challenge and mouse 11, where *C. difficile* was present at a high density from the 4th day post-challenge onward (Fig. S1B). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

```
**5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha diversity continued to increase over time based on the tissues from PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet X). Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)`) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C. difficile*) was one of the genera that had a significant difference in relative abundance between treatment groups at 6 days post-challenge. This population was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). By 30 days post-challenge, only the relative abundances of *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Because there were differences in the mucosal microbiota including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5-2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

```
**Post-challenge laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. To examine the impact of PEG treatment on *C. difficile* clearance, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B).

@@

``` {r ml results}
#AUC results
auc_results <-  read_csv(here("results/performance_results.csv")) %>%
  mutate(taxa_level = "genus") %>%  
  group_by(method) %>%
  summarize(median = median(AUC)) %>% #Median AUC for the 100 seeds of each method
  arrange(desc(median))
#Highest median AUC value out of the 3 methods (random forest) run.
max_auc <- auc_results %>% select(median) %>% max()

```
**Five-day post-challenge community data can predict which mice that will have prolonged *C. difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C. difficile* colonization. We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C. difficile* 10 days post-challenge. We chose 5 days post-challenge because that was the earliest time point where we would see a treatment effect in the post-challenge 1-day PEG experiments. The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C. difficile *colonization. We selected the top 10 bacteria contributing to our model's performance (Fig. 7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C. difficile* colonization status on day 10 (Fig. 7B). Next, we focused on the 5 genera that had a greater than 1 % relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time. We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, the relative abundance of *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C. difficile* (Fig. 7C). Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-challenge 1-day PEG treated mice.

## Discussion
While the disruptive effect of antibiotics on *C. difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear. By following osmotic laxative treated mice over time, we found 5-day PEG treatment before challenge resulted in prolonged *C. difficile* colonization, while a 1-day PEG treatment resulted in transient colonization. The differences in *C. difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota. Additionally, the intestinal communities of 5-day PEG treated mice had not regained colonization resistance after a 10-day recovery period. Although in contrast to the other 5-day PEG treatment groups, *C. difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group. We also examined the impact of PEG treatment after *C. difficile* challenge and in opposition to the hypothesis suggested by the literature, found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment. We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C. difficile* colonization (Fig. 8). Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C. difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after challenge.

In addition to altering composition, laxative treatment may alter microbiota-produced metabolites. A previous study demonstrated that a 5-day treatment of 10% PEG depleted acetate and butyrate and increased succinate compared to untreated mice [@Ferreyra2014]. While we did not perform metabolomic analysis, we did see bacteria known to produce benefical metabolites were depleted in mice that cleared *C. difficile* compared to mice with prolonged colonization. For example, *Oscillibacter valericigenes* can produce the SCFA valerate [@Iino2007], and separate studies demonstrated valerate is depleted after clindamycin treatment and inhibited *C. difficile* growth based on *in vitro* data and an experiment with C57BL/6 mice [@Jump2014; @McDonald2018]. Similarly, *Acetatifactor* can produce acetate and butyrate [@Pfeiffer2012], SCFAs that are decreased in mice with prolonged *C. difficile* infection after antibiotic treatment [@Lawley2012]. Thus protective bacteria and their metabolites could be depleted by osmotive laxative treatment depending on the timing and duration of treatment.
